---
title: "Alcohol Abuse and Mental/Physical Health in the United States"
author: "Jane Condon"
date: "`r Sys.Date()`"
output: 
    pdf_document:
    includes:
      in_header: reformat_paragraph.tex
    df_print: kable
    toc: yes
    toc_depth: 2
    latex_engine: xelatex
    keep_tex: yes
    number_sections: true
editor_options: 
  chunk_output_type: inline

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```



# Loading packages

```{r, message = FALSE, warning = FALSE}
library(haven)
library(sf)
library(ggplot2)
library(sjPlot)
library(sjmisc)
library(sjlabelled)
library(plyr)
library(tidyverse)
library(formatR)
```

# Importing data from the CDC's BRFSS - 2018

```{r}
zip_tf <- tempfile()

zip_url <- "https://www.cdc.gov/brfss/annual_data/2018/files/LLCP2018XPT.zip"
    
download.file( zip_url , zip_tf , mode = 'wb' )

brfss_tbl <- read_xpt( zip_tf )

brfss_df <- data.frame( brfss_tbl )

names( brfss_df ) <- tolower( names( brfss_df ) )

brfss_df[ , 'one' ] <- 1
```


# Data Manipulation and Cleaning


## Selecting only the variable names that we want to keep to make the data more manageable


```{r}
variables <- c("x_state", "maxdrnks", "alcday5", "avedrnk2", "drnk3ge5",
               "menthlth", "poorhlth", "addepev2", "genhlth", "physhlth", "x_ageg5yr", 
               "sex1", "one", "x_llcpwt","x_rfbing5")
brfss_df <- brfss_df[variables]
```



## Adding state abbreviation column

```{r}
fips <- read.csv("fips.csv")

colnames(brfss_df)[colnames(brfss_df) == "x_state"] = "fips"

brfss_df <- merge(fips[,c("fips", "st","stname")], brfss_df, by = "fips", all.x=F, all.y=T)
```


## Subsetting the data


### Binge Drinking Prevalence



```{r}
binge_prev <- brfss_df[(brfss_df$x_rfbing5 == 1 | brfss_df$x_rfbing5 ==2) &
                         !is.na(brfss_df$x_rfbing5),]
```


### Binge Drinking Intensity


```{r}
# Dataframe excluding NA and 'not sure' or 'refused' responses from the 'max drinks' column
max_drinks <- brfss_df[(brfss_df$maxdrnks %in% 1:76) & !is.na(brfss_df$maxdrnks),]
```


### Binge Drinking Frequency


```{r}
# Dataframe excluding NA and 'not sure' or 'refused' responses from the binge drinking column
binge_freq <- brfss_df[(brfss_df$drnk3ge5 %in% 1:88) & !is.na(brfss_df$drnk3ge5),]

#Replacing '88' values with '0', since a response of '88' means 'None'
binge_freq$drnk3ge5[binge_freq$drnk3ge5 == "88"] <- 0
```



### Average Alcohol Consumption

```{r}
# Dataframe excluding NA and 'not sure' or 'refused' responses from the 'average number of drinks' consumed column
avg_drinks <- brfss_df[(brfss_df$avedrnk2 %in% 1:76) & !is.na(brfss_df$avedrnk2),]
```



### Average Drinking Frequency


```{r}
# Dataframe excluding NA and 'not sure' or 'refused' responses from the 'number of days per week/month
# where you consumed an alcoholic beverage' column
alc_days <- brfss_df[(brfss_df$alcday5 %in% 101:107 | brfss_df$alcday5 %in% 201:230 | brfss_df$alcday5 == 888) & !is.na(brfss_df$alcday5),]

# Replacing '888' values with '0', since a response of '888' means 'No drinks in the past 30 days'
alc_days$alcday5[alc_days$alcday5 == "888"] <- 0

# Remove the '2' from 201-230 values, since 2 _ _ = days per month and dividing it by 7 to get 'days per week'

alc_days$alcday5 <- ifelse(alc_days$alcday5 %in% 201:209 ,  gsub("20","",alc_days$alcday5), alc_days$alcday5)

alc_days$alcday5 <- ifelse(alc_days$alcday5 %in% 210:230 ,  gsub("2","",alc_days$alcday5), alc_days$alcday5)

alc_days$alcday5 <- ifelse(alc_days$alcday5 %in% 1:30 ,  round((as.numeric(alc_days$alcday5))/7,2), alc_days$alcday5)

#Remove the '10' from 101-107 values, since 1 _ _ = days per week

alc_days$alcday5 <- ifelse(alc_days$alcday5 %in% 101:107 ,  gsub("10","",alc_days$alcday5), alc_days$alcday5)
```



# Exploratory Data Analysis and Visualizations



## Binge Drinking Prevalence


### Calculating the proportion of binge drinkers in each state


```{r}
yesNo <- data.frame(Yes = rowSums(binge_prev["x_rfbing5"] == 2), No = rowSums(binge_prev["x_rfbing5"] == 1))
binge <- aggregate(yesNo, binge_prev["stname"], sum)
binge$percent <- (binge$Yes)/(binge$Yes + binge$No)*100
binge$percent <- round(binge$percent,2)
```


### Creating a map to display the proportion of binge drinkers in each state


```{r}
states <- st_read("2015-2019-acs-states.geojson")

states <- states[!(states$ST %in% c("PR","AK","HI")),]

binge <- binge[!(binge$stname %in% c("Puerto Rico ","Alaska","Hawaii")), ]

states <- st_transform(states, 6580)

states_prevalence <- merge(states, binge[,c("stname","percent")], by.x = "Name", by.y = "stname")

map <- ggplot() +
    geom_sf(data=states_prevalence, aes(fill=binge$percent)) +
    scale_fill_distiller(palette="PuBuGn", trans = "reverse") +
    labs(title = "Binge Drinking Prevalence \n\ (Prevalence of Binge Drinking Among Adults in the Past 30 Days)", fill = "Prevalence Rate (%)") + theme(plot.title = element_text(hjust = 0.5))

print(map)
```




## Binge Drinking Intensity


### Looking at state means and quartiles of 'most drinks consumed on a single occasion in the past 30 days' among those who engaged in binge drinking in the past 30 days


```{r}
# Creating a subset of the "max drinks" dataframe, only including those who have engaged in binge drinking in the past 30 days (>= 4 drinks for women or >= 5 drinks for men)
max_drinks_binge <- max_drinks[(max_drinks$maxdrnks >= 5 & max_drinks$sex1 == 1) | (max_drinks$maxdrnks >= 4 
                                                                                      & max_drinks$sex1 == 2),]

#Finding state means and quantiles

intensity_state_means <- aggregate(max_drinks_binge$maxdrnks, list(max_drinks_binge$stname), FUN=mean) #Mean
intensity_state_quantiles <- aggregate(max_drinks_binge$maxdrnks, list(max_drinks_binge$stname), FUN=quantile) #Quantiles
intensity_state_75 <- data.frame(intensity_state_quantiles$Group.1,intensity_state_quantiles$x[, "75%"]) #75th percentile

```



### Creating a map to display the state means for binge drinking intensity


```{r}

intensity_state_means <- intensity_state_means[!(intensity_state_means$Group.1 %in% c("Puerto Rico ","Alaska","Hawaii")), ]

states_bngdrnk <- merge(states, intensity_state_means[,c("Group.1","x")], by.x = "Name", by.y = "Group.1")

map <- ggplot() +
    geom_sf(data=states_bngdrnk, aes(fill=x)) +
    scale_fill_distiller(palette="Blues", trans = "reverse") +
    labs(title = "Binge Drinking Intensity \n\ (Max Number of Drinks on a Single Occasion Among Binge Drinkers)", fill = "Number of Drinks") + theme(plot.title = element_text(hjust = 0.5))

print(map)
```



## Binge Drinking Frequency


### Calculating state means for binge drinking frequency among those who engage in binge drinking

```{r}

#Creating a subset of the data, which only includes those who have engaged in binge drinking in the past 30 days 

binge_freq <- binge_freq[(binge_freq$maxdrnks %in% 1:76) & !is.na(binge_freq$maxdrnks),]
binge_freq <- binge_freq[(binge_freq$maxdrnks >= 5 & binge_freq$sex1 == 1) | (binge_freq$maxdrnks >= 4 
                                                                                      & binge_freq$sex1 == 2),]

#Finding state means and quantiles

frequency_state_means <- aggregate(binge_freq$drnk3ge5, list(binge_freq$stname), FUN=mean) #Mean
frequency_state_quantiles <- aggregate(binge_freq$drnk3ge5, list(binge_freq$stname), FUN=quantile) #Quantiles
frequency_state_75 <- data.frame(frequency_state_quantiles$Group.1,intensity_state_quantiles$x[, "75%"]) #75th percentile
```



### Creating a map displaying binge drinking frequency


```{r}
frequency_state_means <- frequency_state_means[!(frequency_state_means$Group.1 %in% c("Puerto Rico ","Alaska","Hawaii")), ]

states_freq <- merge(states, frequency_state_means[,c("Group.1","x")], by.x = "Name", by.y = "Group.1")

map <- ggplot() +
    geom_sf(data=states_freq, aes(fill=x)) +
    scale_fill_distiller(palette="BuPu", trans = "reverse") +
    labs(title = "Binge Drinking Frequency \n\ (Mean Number of Binge Drinking Occasions Among Binge Drinkers)", fill = "Number of Occasions") + theme(plot.title = element_text(hjust = 0.5))

print(map)
```



## Depressive Disorders and Mental/Physical Health Condition vs Binge Drinking 


```{r}
# Creating a subset of the data, excluding NA or 'not sure'/'refused' values for binge drinking, depressive disorder, physical health, and general health

depression_prev <- binge_prev[(binge_prev$addepev2 ==1 | binge_prev$addepev2 ==2) & !is.na(binge_prev$addepev2),]
health_prev <- depression_prev[(depression_prev$genhlth %in% 1:5) & !is.na(depression_prev$genhlth) & 
                                 (depression_prev$menthlth %in% 1:30 | depression_prev$menthlth ==88) & 
                                 !is.na(depression_prev$menthlth) & (depression_prev$physhlth %in% 1:30 | depression_prev$physhlth ==88) & !is.na(depression_prev$physhlth),]
health_prev$menthlth <- as.numeric(gsub("88","0", health_prev$menthlth)) # Changing '88' to '0', since '88' = 'None'
health_prev$physhlth <- as.numeric(gsub("88","0", health_prev$physhlth)) # Changing '88' to '0', since '88' = 'None'
```


### Constructing a Logistic Regression Model to Predict Whether a Survey Respondent is a Binge Drinker Based on Their Health Condition


### Preparing the Data for Logistic Regression 


```{r}

health_prev$x_rfbing5 <- as.numeric(gsub("1","0",health_prev$x_rfbing5)) # No = 0
health_prev$x_rfbing5 <- as.numeric(gsub("2","1",health_prev$x_rfbing5)) # Yes = 1

health_prev$addepev2 <- as.numeric(gsub("2","0",health_prev$addepev2)) # No = 0, Yes = 1

health_prev <- health_prev[c("x_rfbing5","addepev2","menthlth","physhlth","genhlth")]


col_names <- c("x_rfbing5","addepev2")

health_prev[,col_names] <- lapply(health_prev[,col_names] , factor)


```




### Logistic Regression Model: Binge Drinker ~ Depressive Disorder + Mental Health + Physical Health + General Health


```{r}
model <- glm(x_rfbing5 ~ addepev2 + menthlth + physhlth + genhlth, family=binomial(link='logit'),data=health_prev)
summary(model)
```



```{r, results = 'hide'}
#Displaying results of logistic regression as a table
tab_model(model, dv.labels = "Binge Drinker (No = 0, Yes = 1)", pred.labels = c("(Intercept)","Depressive Disorder", "Mental Health", "Physical Health","General Health"), show.est = TRUE, show.stat = TRUE)
```



### Stacked Bar Plot Showing Binge Drinking vs. Deppressive Disorder



```{r}
dat <- data.frame(table(health_prev$addepev2,health_prev$x_rfbing5))
names(dat) <- c("addepev2","x_rfbing5","count")
dat <- ddply(dat, .(addepev2), transform, percent = count/sum(count) * 100)
dat$label = paste0(sprintf("%.0f", dat$percent), "%")

ggplot(data=dat, aes(x=addepev2, y=count, fill=x_rfbing5)) + geom_bar(stat="identity")  + scale_x_discrete(labels=c('No','Yes')) + scale_fill_manual(values = c("#99bbde","#06478a"), labels = c("No","Yes"))  + labs(title = "Binge Drinking vs. Depressive Disorder", fill = "Binge Drinker", x = "Diagnosed with a Depressive Disorder", y = "Count") + geom_text(aes(label = label), hjust = 0.5, vjust = 1, size = 4,color = "white") +
  theme(plot.title = element_text(hjust = 0.5)) + scale_y_continuous(labels = scales::comma_format())
```



### Binge Drinking vs. Poor Mental and Physical Health



```{r}
dat2 <- health_prev[c("x_rfbing5","menthlth","physhlth")]
dat2 <- aggregate(. ~ x_rfbing5, dat2, mean)

dat2$rfbing5 <- row.names(dat2)
dat2_long <- gather(dat2, key = var, value = value, menthlth, physhlth)

ggplot(dat2_long, aes(x = x_rfbing5, y = value, fill = var)) +
  geom_bar(stat = 'identity', position = 'dodge') + scale_x_discrete(labels=c('No','Yes')) + scale_fill_manual(values = c("#b4b4d1","#3c2b59"), labels = c("Mental Health","Physical Health")) + labs(title = "Binge Drinking vs. Mental and Physical Health", fill = "Type of Health",x = "Binge Drinker", y = "Number of Poor  Health Days Per Month (Mean)") + geom_text(aes(label = round(value,2)), vjust = 1.5,
             position = position_dodge(.9), size = 4, color = "white") +
  theme(plot.title = element_text(hjust = 0.5))
```



### Binge Drinking vs. General Health


```{r}
dat3 <- aggregate(genhlth ~ x_rfbing5, health_prev, median)
ggplot(data=dat3, aes(x=x_rfbing5, y=genhlth, fill = x_rfbing5)) +
  geom_bar(stat="identity", width=0.5) + scale_x_discrete(labels=c('No','Yes')) + scale_fill_manual(values = c("#a7d4bf","#3a6953"), labels = c("No","Yes")) + labs(title = "Binge Drinking vs. General Health", fill = "Binge Drinker",x = "Binge Drinker", y = "Median General Health Rating (1-5)") + geom_text(aes(label = round(genhlth,2)), hjust = 0.5, vjust = 4, size = 4,color = "white") +
  theme(plot.title = element_text(hjust = 0.5))
```

